---
title: "Updating R and RStudio"
author: "Nick Hathaway; nicholas.hathaway@umassmed.edu"
output:
  html_document:
    highlight: tango
    theme: readable
    toc: true
  pdf_document:
    fig_caption: yes
    highlight: tango
    toc: true
---

```{r setup, echo=FALSE, message=FALSE}
require(knitr)
#turn off mesages and warnings and make it so output isn't prefixed by anything, default is to put "##" in front of all output for some reason
opts_chunk$set(message=FALSE, warning=FALSE, comment = "")
```

#Updating R

##Updating on Windows
Updating on Windows is apparently tricky, a package called `installr`, which is only for Windows can be used to ease the process.  
The following commands were taken from [here](http://www.r-statistics.com/2013/03/updating-r-from-r-on-windows-using-the-installr-package/), which goes into more details.  
First Install the `installr` package if you don't have it
```{r, eval = F}
# installing/loading the package:
if(!require(installr)) {
  install.packages("installr"); 
  require(installr)
} #load / install+load installr
```
Now call `updateR()` function to call update  
This will start the updating process of your R installation.  It will check for newer versions, and if one is available, will guide you through the decisions you'd need to make.
```{r, eval = F}
# using the package:
updateR()
```
##Updating On Mac and Ubuntu
One Mac and Ubuntu you need to go to the [CRAN](https://www.r-project.org/) website, https://www.r-project.org/ to install the newer package intaller.

#Updating RStudio
Updating RStudio is easy, just go to `Help` > `Check for Updates` to install newer version.  

#Updating R Packages

##Updating Out of date Packages
Updating out of date package that were installed with `install.packages()` is easy with the `update.packages()` function.
```{r, eval = F}
update.packages()
```
Which will ask you for every package if you want to update, to just say yes to everything use `ask = FAlSE`. 
```{r, eval = F}
update.packages(ask = FALSE)
```
Unfortunately this won't update packages installed by `devtools::install_github()`

##Updating all Packages after R update
On Windows this is hanlded by the package `installr`. But for other operating systems, after updating R, a lot of packages might have been build under the old R version and it would be safer to re-install all the packages already installed. This can be done with the following commands. Again this won't update packages installed with `devtools::install_github()`  
```{r, eval = F}
## get packages installed
packs = as.data.frame(installed.packages(.libPaths()[1]), stringsAsFactors = F)

## and now re-install install packages using install.packages()
install.packages(packs$Package)

```


