---
title: "Session 3"
author: "Nick Hathaway; nicholas.hathaway@umassmed.edu"
output:
  html_document:
    highlight: tango
    theme: readable
    toc: yes
    toc_float: yes
  pdf_document:
    highlight: tango
    toc: yes
---

<!--Include script for hiding output chunks-->
<script>
$(document).ready(function() {
	console.log("hello")
  // Select all <pre> tags that do not have class 'r'
  $output = $(".toggleOutput");
  console.log($output)
  // Add the show/hide-button to each output chunk
  $output.prepend("<button style=\"float:right\" class=\"btn btn-primary showopt\">Show Output</button><br/>");
  // Select the <pre> tags, then choose their <code> child tags and toggle visibility 
  $output.children("code").css({display: "none"});
  
  // When the show/hide-button is clicked, toggle the current state and
  // change the button text
  $(".showopt").click(function() {
    $btn = $(this);
    $chunk = $(this).parent().children("code");
    if($btn.html() === "Show Output") {
      $btn.html("Hide Output");
    } else {
      $btn.html("Show Output");
    }
    $btn.toggleClass("btn-info btn-primary");
    $chunk.slideToggle("fast", "swing");
  });
});

</script>





```{r setup, echo=FALSE, message=FALSE}
require(knitr)
#turn off mesages and warnings and make it so output isn't prefixed by anything,
#default is to put "##" in front of all output for some reason
#also set tidy to true so code is wrapped properly 
opts_chunk$set(message=FALSE, warning=FALSE, comment = "")
options(width = 60)

knitr::knit_hooks$set(output = function(x, options) {
  return(paste0(
    "```{",
    ifelse(is.null(options$class),
      "", 
      paste0(" .", gsub(" ", " .", options$class))
    ),
    "}\n",
    x,
    "\n```"
  ))
})
```

# Part 1. Excercises Solutions

Download Average Temperatures USA  
<a download = "avg_temps_usa_wide.tab.txt" href = "avg_temps_usa_wide.tab.txt">Average Temperatures USA</a>  

1.  convert to **long** format by using gather() on the temperate columns    
2.  separate columns so you have a column for year and month  
3.  create a table of mean temperatures for months, cities, and months for each city (various group_by calls)  
4.  filter table to just one city or just one month  

# Part 1 Solutions 

## Q1

convert to **long** format by using gather() on the temperate columns    

```{r}
library(tidyverse)  
temps = readr::read_tsv("avg_temps_usa_wide.tab.txt")
# temperature columns start on column 6 
temps_gat = temps %>% 
  gather(date, temperature, 6:ncol(temps))

```

## Q2 

separate columns so you have a column for year and month  

```{r}
# Separate and then convert year 
temps_gat = temps_gat %>% 
  separate(date, into = c("year", "month")) %>% 
  mutate(year = as.numeric(year))

### OR

# separate and let separate guess the type of the separated columns use convert = T 
temps = readr::read_tsv("avg_temps_usa_wide.tab.txt")
# temperature columns start on column 6 
temps_gat = temps %>% 
  gather(date, temperature, 6:ncol(temps))
temps_gat = temps_gat %>% 
  separate(date, into = c("year", "month"), convert = T)


```

## Q3  
create a table of mean temperatures for months, cities, and months for each city (various group_by calls)  

```{r}
# filter off NA 
temps_gat_filt = temps_gat %>% 
  filter(!is.na(temperature))

## Summary within cities 
temps_gat_summary_city = temps_gat_filt %>% 
  group_by(Station_Name) %>% 
  summarise(meanTemp = mean(temperature))

## Summary within months
temps_gat_summary_month = temps_gat_filt %>% 
  group_by(month) %>% 
  summarise(meanTemp = mean(temperature))

## Summary of within months within each city 
temps_gat_summary_city_month = temps_gat_filt %>% 
  group_by(Station_Name, month) %>% 
  summarise(meanTemp = mean(temperature))


```

## Q4  
filter table to just one city or just one month 

```{r}
# Filter to just one city, here BURLINGTON/INT'L ARPT
temps_gat_BURLINGTON = temps_gat_filt %>% 
  filter(Station_Name == "BURLINGTON/INT'L ARPT")

# Filter to just one year, here 1994
temps_gat_1994 = temps_gat_filt %>% 
  filter(year == 1994)

```


# Part 2. Excercises

Using the Temperature data frame read in earlier  
<a download = "avg_temps_usa_wide.tab.txt" href = "avg_temps_usa_wide.tab.txt">Average Temperatures USA</a>  

1. Filter the long format data frame created in Part 1 to just one Station_name  
2. Modify the month column into a factor so that the months are organized in chronological order. (hint use this vector `c("January","February","March","April","May","June","July","August","September","October","November","December"))`  
3. Create a line and dot plot of temperature for the Station_name you picked in 1 with months on x-axis and temperatures on y-axis, color the lines by years (see what happens to the colors when you change years into a factor rather than a numeric data type)   
4. Now create a barplot 
5. Now filter the long format data frame again to be from 3 different stations and to just the year 1995. 
6. Take the new data frame from 5 and create a barplot x = months and y = temperature and color the bars by station name (try setting the station names to new custom colors of your choosing, you can use http://colorbrewer2.org/ to pick colors)

# Part 2 Solutions 

## Q1

Filter the long format data frame created in Part 1 to just one Station_name

```{r}
temps = readr::read_tsv("avg_temps_usa_wide.tab.txt")
# temperature columns start on column 6 
temps_gat = temps %>% 
  gather(date, temperature, 6:ncol(temps))
temps_gat = temps_gat %>% 
  separate(date, into = c("year", "month"), convert = T)
# filter off NA 
temps_gat_filt = temps_gat %>% 
  filter(!is.na(temperature))
# Filter to just one city
temps_gat_boston = temps_gat_filt %>% 
  filter(Station_Name == "BOSTON/LOGAN INT'L ARPT")
```

## Q2  
Modify the month column into a factor so that the months are organized in chronological order. (hint use this vector `c("January","February","March","April","May","June","July","August","September","October","November","December"))`  
```{r}
temps_gat_boston = temps_gat_boston %>% 
  mutate(month = factor(month, levels = c("January","February","March","April","May","June","July","August","September","October","November","December")))

```

## Q3    
Create a line and dot plot of temperature for the Station_name you picked in 1 with months on x-axis and temperatures on y-axis, color the lines by years (see what happens to the colors when you change years into a factor rather than a numeric data type)   

theme(axis.text.x = element_text(angle = -45, hjust = 0)) allows the rotation of the month labels so they fit better 
```{r}
linePlot = ggplot(temps_gat_boston,aes(x = month, y = temperature, color = factor(year), group = factor(year)  )) + 
  geom_line() +
  geom_point() + 
  scale_color_brewer(palette = "Paired") + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) + 
  labs(x = "Month", y = "Temperature (C)", color = "Year")

print(linePlot)
```

## Q4  
Now create a barplot  
```{r}
barPlot = ggplot(temps_gat_boston) + 
  geom_bar(aes(x = month, y = temperature, fill = factor(year), group = factor(year)  ), stat = "identity", position = "Dodge") + 
  scale_fill_brewer(palette = "Paired") + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) + 
  labs(x = "Month", y = "Temperature (C)", fill = "Year")

print(barPlot)

```


## Q5  
Now filter the long format data frame again to be from 3 different stations and to just the year 1995.

```{r}
# Filter to just three city
stations = c("BOSTON/LOGAN INT'L ARPT", "TUCSON/INT'L ARPT", "OLYMPIA/ARPT")
temps_gat_threeStats_1995 = temps_gat_filt %>% 
  filter(Station_Name %in%  stations, 
         year == 1995)

temps_gat_threeStats_1995 = temps_gat_threeStats_1995 %>% 
  mutate(month = factor(month, levels = c("January","February","March","April","May","June","July","August","September","October","November","December")))

```

## Q6  
Take the new data frame from 5 and create a barplot x = months and y = temperature and color the bars by station name (try setting the station names to new custom colors of your choosing, you can use http://colorbrewer2.org/ to pick colors)  
```{r}
stationColors = c('#e41a1c','#377eb8','#4daf4a')
names(stationColors) = stations

barPlot = ggplot(temps_gat_threeStats_1995) + 
  geom_bar(aes(x = month, y = temperature, fill = Station_Name, group = Station_Name ), stat = "identity", position = "Dodge", color =  "black") + 
  scale_fill_manual(values = stationColors) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0), 
        legend.position = "bottom", 
        panel.border = element_blank()) + 
  labs(x = "Month", y = "Temperature (C)", fill = "Station") 

print(barPlot)

```



