---
title: "Ensuring A Library/Package Is Installed"
author: "Nicholas Hathaway"
output:
  html_document:
    toc: yes
    highlight: tango
    theme: readable
  pdf_document:
    toc: yes
    highlight: tango
---

```{r setup, echo=FALSE, message=FALSE}
require(knitr)
#turn off mesages and warnings and make it so output isn't prefixed by anything,
#default is to put "##" in front of all output for some reason
#also set tidy to true so code is wrapped properly 
opts_chunk$set(message=FALSE, warning=FALSE, comment = "")
options(width = 60)
```

There are several ways of loading a package, two of these ways are `library()` and `require()` functions. Both of these function load the functions that are in the requested library but they return two separate things. 

##library()
The `library()` function actually returns all the libraries currently loaded when you call it
```{r, eval=T}
libs = library(dplyr)
print(libs)
```

##require()
The `require()` function returns whether or not the library was able to be loaded, which if it fails is normally due to the fact that it doesn't exist
```{r, eval=T}
libLoaded = require(dplyr)
print(libLoaded)
```

That means you can take advantage of the return of `require()` to ensure a packaged is installed without having to call `install.packages()` every time you want a package. 
```{r, eval=T}
if(!require(dplyr)){
  install.packages("dplyr")
  #call require again because install.packages simply installs the package, it doesn't load it
  require(dplyr)
}
```
If you put the above code at the top of your script it will only call `install.package` if it has to. You can also turn the above into a function to be reused. 
```{r, eval=T}
ensureLib <-function(libName){
  if(!require(libName, character.only = T)){
    install.packages(libName)
    #call require again because install.packages simply installs the package, it doesn't load it
    require(libName, character.only = T)
  }
}

ensureLib("dplyr")
```


