---
title: 'Data Visualization in R: Session 4'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
date: "October 8, 2014"
---
```{r setup, echo=FALSE, message=FALSE}
require(knitr)
library(png)
library(grid)
#turn off mesages and warnings and make it so output isn't prefixed by anything,
#default is to put "##" in front of all output for some reason
#also set tidy to true so code is wrapped properly 
#opts_chunk$set(message=FALSE, warning=FALSE, comment = "", tidy = T)
opts_chunk$set(message=FALSE, warning=FALSE, comment = "", tidy=TRUE)
options(width = 60)
#options(rpubs.upload.method = "internal")

options(RCurlOptions = list(verbose = FALSE, capath = system.file("CurlSSL", "cacert.pem", package = "RCurl"), ssl.verifypeer = FALSE))

```

************

#Part 1 Installing and loading R packages
In this session we will use an R package called `ggplot2` to provide us with easiers ways to create complex plots. But first we have to learn how to download the packages which can be done using `install.packages()` and `library()` or we can use RStudio to download them. 

Using `install.packages()` and `library()`:
```{r, eval=FALSE}
#install package
install.packages("ggplot2")
#Once package is installed to actually use it you need to load it
library("ggplot2")
```  

Or we can use RStudio to manually install the packages. In the bottom right corner of RStudio click the Packages tab and then the install icon.   
```{r fig.width=8, echo=FALSE}
grid.raster(readPNG("../images/installRPackage_1.png"))
```  

Type in ggplot2  
```{r fig.width=4, echo=FALSE}
grid.raster(readPNG("../images/install_ggplot.png"))
```  

To load click the check box next to ggplot2  

```{r fig.width=8, echo=FALSE}
grid.raster(readPNG("../images/install_ggplot2.png"))
```  

You can see the output in the console window when doing this by hand and it actually is doing what we did before by calling `install.packages()` and `library()`:  

```{r fig.width=8, echo=FALSE}
grid.raster(readPNG("../images/install_ggplot3.png"))
```  

# Part 2: Plotting using ggplot2

## Part 2. Download data
We are using the dataset from last session; frequency of several diseases in Boston since the 1880's [BosEpi.tab.txt](http://bib.umassmed.edu/~hathawan/rCourse/dataSets/BosEpi.tab.txt). Download and read the data into R.

```{r, eval=FALSE}
bosEpi <- read.delim(file.choose())
```

```{r, echo=FALSE}
bosEpi <- read.delim("../dataSets/BosEpi.tab.txt")
worEpi <- read.delim("../dataSets/WorEpi.tab.txt")
```

View data table 
```{r, eval=FALSE}
View(bosEpi)
```

`qplot()` is a superior alternative to `plot()`. It is part of the "ggplot2" package and provides easy ways to produce publication-quality figures. For example:  

```{r}
library(ggplot2) # be sure to load the "ggplot2" library!
```

```{r fig.width=10}
qplot(Year, number, geom="histogram", stat="identity", fill=disease, position="dodge", data=bosEpi[bosEpi$event == "DEATHS",], xlim=c(1885, 1940))
```

# Part 3: step-by-step qplot()  
We will now start with a few small examples of `qplot()`.

First, let's generate some random data.
```{r}
randomData <- rnorm(10000) # normally distributed random data
```

The default for `qplot()` is to produce a histogram.
```{r}
qplot(randomData)
```

## Part 3. Boston disease data
First, let's extract the data and sum it again, like in the last session.  
```{r}
dipScarDeaths <- bosEpi[(bosEpi$disease == "DIPHTHERIA" | bosEpi$disease == "SCARLET FEVER") & bosEpi$event == "DEATHS",]
dipScarDeathsSum <- aggregate(number~disease*Year, dipScarDeaths, sum)
head(dipScarDeathsSum)
```

Now, this is how we can use `qplot()` to plot the number of deaths per year, using the  aggregated data.  
```{r}
qplot(Year, number, data=dipScarDeathsSum)
```

In order to see which points correspond to which disease, we can color by disease.  
```{r}
qplot(Year, number, data=dipScarDeathsSum, color=disease)
```

Also, we may not be interested in the diphtheria death outlier in 1980, so, let's change the scale of the x-axis.  
```{r}
qplot(Year, number, data=dipScarDeathsSum, color=disease, xlim=c(1880, 1930))
```

What we see is that `qplot()` automatically chose reasonable colors and also produced a legend.  

To turn this plot into a bar plot, we can do the following.  
```{r, fig.width=10}
qplot(Year, number, data=dipScarDeathsSum, color=disease, xlim=c(1880, 1930), geom="bar", stat="identity", position="dodge", fill=disease, main = "Deaths by Diphtheria vs Deaths by Scarlet Fever", xlab = "Year", ylab = "Number of Deaths")
```

The important parts of this command are:  

*  **data** specifies the data frame to use  
*  **Year** and **number** are columns from the data set  
*  **color** sets the color of the border of the bars, in this case it is based on the "disease" column  
*  **xlim** limits the scale of the x-axis, like in `plot()`  
*  **xlab**, **ylab** and **main** are titles, like in `plot()`  

To illustrate the difference between `plot()` and `qplot()`, below are the commands that were used last week to generate a similar figure. `qplot()` make this much simpler.  
```{r}
plot(dipScarDeathsSum[dipScarDeathsSum$disease == "SCARLET FEVER",]$Year,
     dipScarDeathsSum[dipScarDeathsSum$disease == "SCARLET FEVER",]$number,
     type = "h", xlab = "Year", ylab = "Number of Deaths", 
     main = "Deaths by Diphtheria vs Deaths by Scarlet Fever",
     bty = "none", lwd = 2, xlim=c(1880, 1930), 
     ylim = range(dipScarDeathsSum$number), col= "red")  

points(dipScarDeathsSum[dipScarDeathsSum$disease == "DIPHTHERIA",]$Year + 0.5,
       dipScarDeathsSum[dipScarDeathsSum$disease == "DIPHTHERIA",]$number,
       type = "h", lwd = 2, col = "darkblue") 

legend("topright", legend = c("SCARLET FEVER", "DIPHTHERIA"),
       lwd = 2, col = c("red", "darkblue"))
```

##Part 3 Exercises {#part3_questions}
1.  Try customizing some plots for the diseases found in the [Worcester Epi Dataset](http://bib.umassmed.edu/~hathawan/rCourse/dataSets/WorEpi.tab.txt)  
2.  For example trying to look how deaths have decreased over time for some diseases like measles (though maybe on the raise again), smallpox, etc.
3.  Try creating a plot comparing deaths between "SCARLET FEVER"" and "TYPHOID FEVER [ENTERIC FEVER]"
4.  Try ploting a couple of the big diseases in the 1920's 

#Part 4 Complex plots on large data
Here's the link to the data set you will be using for the next exercises. This dataset contains disease information from  major cities in the US.  

[someEpi.tab.txt](http://bib.umassmed.edu/~hathawan/rCourse/dataSets/someEpi.tab.txt)  

You can download the file and then read it into R using `read.delim()`. Also, you can read it directly, like this:

```{r}
epi <- read.delim("http://bib.umassmed.edu/~hathawan/rCourse/dataSets/someEpi.tab.txt")
head(epi)
```

```{r}
x <- aggregate(number ~ disease * month * event, epi[epi$disease == "INFLUENZA", ], sum)
qplot(month, number, data=x, geom="path", facets=~event)
```

```{r}
x <- aggregate(number ~ state * disease * month * event, epi[epi$disease == "INFLUENZA", ], sum)
qplot(month, number, data=x, geom="path", group=state, color=state, facets=~event) + guides(col = guide_legend(ncol=3))
```

```{r}
x <- aggregate(number ~ state * disease * month * event, epi[epi$disease == "INFLUENZA", ], sum)
qplot(month, number, data=x, geom="smooth", facets=~event)
```

##Part 4. Exercises.
1.  Pick a Nothern and a Southern state. Is there a noticable difference in the pattern of influenza cases and deaths?
2.  Pick a couple of diseases and plot graphs for the cities chosen.
3.  Facet the graphs by event and disease, i.e. look at whooping cough vs. measles.

```{r fig.height=8}
library(scales)
qplot(disease, number, data=aggregate(number ~ disease, epi, sum), geom="bar", stat="identity", main="Total events per disease") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + scale_y_continuous(labels = comma)
```

