---
title: "Additional Plotting Aspects"
output: 
  html_document
---
```{r setup, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, comment = "")
```
# Changing dot type, dot size, and line type
Several customizable plotting options include 

*  Changing the point shape using `pch=` (options are 0 through 25)  
*  Changing the point size using `cex=`  
*  Changing the line type for lines using `lty=` (options are 0 through 6)  

Example for changing shape    
There are 26 available shapes which can be specified (default is 1 a hollow dot)    
```{r}
plot(0:25, pch = 0:25)
```

Also for the shapes 21-25 the `col=` parameter changes the border color and the `bg=` changes the inner color  

```{r}
plot(0:25, pch = 0:25, col = "red", bg = "darkblue")
```

You can also change the size of dots by using `cex=`.  You give it a relative value, something that has a `cex` value of 4 will be twice as big as something that has a `cex`  value of 2, but the number itself doesn't mean a size

```{r}
plot(1:10, cex = 1:10)
```

When doing line plots, you can change the type of line being drawn by using `lty=`  There are seven options given by giving lty 0 through 6 

number | type of line  
-------|--------  
0|blank  
1|solid  
2|dashed  
3|dotted  
4|dotdash
5|longdash
6|twodash

```{r, eval=FALSE}
# blank you can't see it
plot(c(0,1), c(0,0), ylim = c(0,6), type="l", lty=0)
# solid (the default)
lines(c(0,1), c(1,1), type="l", lty=1)
# dashed
lines(c(0,1), c(2,2), type="l", lty=2)
# dotted
lines(c(0,1), c(3,3), type="l", lty=3)
# dotdash
lines(c(0,1), c(4,4), type="l", lty=4)
# longdash
lines(c(0,1), c(5,5), type="l", lty=5)
# twodash
lines(c(0,1), c(6,6), type="l", lty=6)
```

```{r, echo=FALSE}
# blank you can't see it
plot(c(0,1), c(0,0), ylim = c(0,6),xlim = c(0,2), type="l", xlab = NA, ylab = NA, bty = "none", lty=0, xaxt = "none")
# solid (the default)
lines(c(0,1), c(1,1), type="l", lty=1)
# dashed
lines(c(0,1), c(2,2), type="l", lty=2)
# dotted
lines(c(0,1), c(3,3), type="l", lty=3)
# dotdash
lines(c(0,1), c(4,4), type="l", lty=4)
# longdash
lines(c(0,1), c(5,5), type="l", lty=5)
# twodash
lines(c(0,1), c(6,6), type="l", lty=6)
# blank you can't see it
text(1.25,0, "blank    (lty = 0)")
text(1.25,1, "solid    (lty = 1)")
text(1.25,2, "dashed   (lty = 2)")
text(1.25,3, "dotted   (lty = 3)")
text(1.25,4, "dotdash  (lty = 4)")
text(1.25,5, "longdash (lty = 5)")
text(1.25,6, "twodash  (lty = 6)")
```

# Using logic to help with plotting  

A helpful to add to a plot is the `ifelse()` function.  You give it three things

1.  A vector of TRUEs and FALSEs  
2.  A value for TRUE  
3.  A value for FALSE  

What this function does is return a new vector by replacing wherever there are TRUEs in the input with the value for TRUE and wherever there are FALSEs with the value for FALSE 

For example  
```{r}
ifelse(c(TRUE,FALSE,TRUE, FALSE), "Replace_TRUE", "Replace_FALSE")
```

This can also work by giving it a logic test on a vector (which will be converted into a vector of TRUEs and FALSEs anyways)
```{r}

testNumbers = c(1,2,3,4)

testNumbers > 2

ifelse(testNumbers >2, "Greater_than_two", "Less_than_two")

```

And you can of course combine logic tests  
```{r}
ifelse(testNumbers >2 & testNumbers < 4, "Greater_than_two_less_than_four", "Less_than_two_or_greater_than_four")
```
 
This function can be helpful for change colors or shapes of scatter plots to demonstrate thresholds 

For example lets look at a scatter plot of 100 randomly generated points with x ranging from -1 to 1 and y ranging from -1 to 1 

```{r}
testNumberData = data.frame(x = runif(100,-1,1), y = runif(100,-1,1))

plot(testNumberData$x, testNumberData$y)

# add a line for x = 0 and y = 0

abline(v = 0)

abline(h = 0)
```

If you were interested in marking points that have a positive x value you could use `ifelse()` with `col` along with the package RColorBrewer to color the points differently 

```{r, eval=FALSE}
install.packages("RColorBrewer")
```

```{r}
library(RColorBrewer)
cols = brewer.pal(2, "Dark2")

plot(testNumberData$x, testNumberData$y, 
     col = ifelse(testNumberData$x > 0, cols[1], cols[2]))

# add a line for x = 0 and y = 0

abline(v = 0)

abline(h = 0)
```

And if you were interested in marking points that had both a positive x value and a positive y value you could use `ifelse()` like this  

```{r}
cols = brewer.pal(2, "Dark2")

plot(testNumberData$x, testNumberData$y, 
     col = ifelse(testNumberData$x > 0 & testNumberData$y > 0,
                  cols[1], cols[2]))

# add a line for x = 0 and y = 0

abline(v = 0)

abline(h = 0)
```
You could also use it to change the shapes as well to mark it even further.  Giving hollow dots (pch = 1) to any points with negative values and solid filled in points (pch=16) to points with only positive values

```{r}
cols = brewer.pal(2, "Dark2")

plot(testNumberData$x, testNumberData$y, 
     col = ifelse(testNumberData$x > 0 & testNumberData$y > 0,
                  cols[1], cols[2]),
     pch = ifelse(testNumberData$x > 0 & testNumberData$y > 0,
                  16, 1))

# add a line for x = 0 and y = 0

abline(v = 0)

abline(h = 0)
```

Now things might get a little messy with too many values if we had a bigger dataset like bellow 

```{r}
cols = brewer.pal(2, "Dark2")

testNumberDataLarge = data.frame(x = runif(2000,-1,1), y = runif(2000,-1,1))

plot(testNumberDataLarge$x, testNumberDataLarge$y, 
     col = ifelse(testNumberDataLarge$x > 0 & testNumberDataLarge$y > 0,
                  cols[1], cols[2]),
     pch = ifelse(testNumberDataLarge$x > 0 & testNumberDataLarge$y > 0,
                  16, 1))

# add a line for x = 0 and y = 0

abline(v = 0)

abline(h = 0)
```

Using the package scales and the `alpha()` function we can make the colors transparent

```{r, eval=FALSE}
install.packages("scales")
```

```{r}
library(scales)

cols = alpha(brewer.pal(2, "Dark2"), alpha = .50)

testNumberDataLarge = data.frame(x = runif(2000,-1,1), y = runif(2000,-1,1))

plot(testNumberDataLarge$x, testNumberDataLarge$y, 
     col = ifelse(testNumberDataLarge$x > 0 & testNumberDataLarge$y > 0,
                  cols[1], cols[2]),
     pch = ifelse(testNumberDataLarge$x > 0 & testNumberDataLarge$y > 0,
                  16, 1))

# add a line for x = 0 and y = 0

abline(v = 0)

abline(h = 0)
```

