---
title: 'Session 5: ggplot2 - The basics'
author: "Adam Wespiser"
date: "November 4, 2014"
output:
  html_document:
    toc: yes
    highlight: tango
    theme: readable
  pdf_document:
    toc: yes
    highlight: tango
---
```{r setup, echo=FALSE, message=FALSE}
require(knitr)
#turn off mesages and warnings and make it so output isn't prefixed by anything,
#default is to put "##" in front of all output for some reason
opts_chunk$set(message=FALSE, warning=FALSE, comment = "")
options(width = 60)
```
## Intro
This course includes both example code and exercises for you to learn the concepts.   
Before starting, please make sure the following libraries are loaded. If you cannot load a  
library, you probably don't have it, and should use the  `install.packages()` function to 
load them.  

## ggplot2 basics
ggplot2 was devloped by Hadley Wickham, a notable R programmer and contributor to Rstudio.  
The project stable, and should be available for the long term.  
Use http://docs.ggplot2.org/current/ as a reference.  
http://www.r-bloggers.com/ is a collection of recent articles on R.  
To save a plot, use `ggsave`  



```{r}
#source("http://bib.umassmed.edu/~wespisea/rCourse/adamGgplot2/answers.R")
library(ggplot2)
library(plyr)
library(grid)
library(reshape2)
library(gridExtra)
library(MASS)
library(HistData)
```

Load the datasets...
```{r, echo=FALSE,cache.lazy=TRUE}
boston.epi.df <- read.csv(file = "http://bib.umassmed.edu/~hathawan/rCourse/dataSets/BosEpi.tab.txt",
                          stringsAsFactors=FALSE,
                          sep="\t")
```

\pagebreak  



##Scatter Plots
To create a scatter plot using qplot you would use:
```{r,eval=FALSE}
qplot(data=iris,x=Sepal.Width,y=Sepal.Length,geom="point")
```
For ggplot2, the command is:  

```{r,fig.width=4,fig.height=3,cache.lazy=TRUE}
ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length)) + geom_point() 
```  

### Question 1
Use the women dataset to plot average heights and weights for american women.  
First, check the contents of the dataset using `str(women)` and use 
`help(women)` to learn about the data. 
Check the answer by `cat(answer1)`.  

\pagebreak  


## Aesthetic Mappings
we can map: color, shape to data.frames or arbritrary values
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length,shape=Species)) + geom_point()
```

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length,color=Species)) + geom_point()
```  


### Question 2
 Using the `birthwt` dataset, what is the distribution of age vs. weight?  
By coloring the points, is low birth weight associated with either variable?  
Check the answer by `cat(answer2)`.  


\pagebreak  




#### Set aesthetics to fixed values:

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length)) + geom_point(color="red",size=4)  
``` 

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(diamonds,aes(x=carat,y=price))+ geom_point(alpha=I(0.4))
```  

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(diamonds,aes(x=carat,y=price))+ geom_point(alpha=I(0.1))
```  


#### Layered geoms  

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}

ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length,color=Species,shape=Species)) + 
  geom_point(size=4) + 
  geom_point(size=2,color="grey")
```   

\pagebreak  


## Line Plot
note, the aesthetic mappings work the same
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(economics,aes(x=date,y=uempmed))+geom_line()
```

### Question 3 
plot the number of Typhoid Fever deaths per year using the epi dataset. 
If you are stuck, take a look at `cat(hint3)` and `cat(hint3.1)` for the data transforms. 
Check the answer by `cat(answer3)`.  

#### Add labels and a title:

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}

ggplot(economics,aes(x=date,y=uempmed))+ geom_line() +
  xlab("Time") + 
  ylab("Unemployment Rate") + 
  ggtitle("Unemployment Rate vs Time\n1967-2007")
```

### Question 4 
Add labels and a title to the previous plot. 
Example solution: `cat(answer4)`.  
\pagebreak  


## Bar Plot
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(mtcars,aes(x=cyl))+geom_bar(binwidth=1) 
```  


#### use factor instead of numerical datatypes
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(mtcars,aes(x=factor(cyl)))+geom_bar()
```  

####color the bars by type
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(mtcars,aes(x=factor(cyl),fill=factor(gear)))+geom_bar()
````

####another, better example:
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(diamonds, aes(clarity, fill=cut)) + geom_bar()
```

## Question 5
What is the distribution of cases of typhoid per week? What about per year?  
See `cat(answer5)` and `cat(answer5.1)`  

\pagebreak  



### Density Plot
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(movies, aes(length))+geom_density()
```

#### Adjust x and y limits
we need to "zoom in" to get a resonable view
create distributions of short vs. non-short movies:
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(movies, aes(x=length,fill=factor(Short)))+geom_density() + xlim(0,200)
```

#### Set transperancy of distro
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(iris, aes(x=Sepal.Width,fill=Species))+geom_density(alpha=I(0.4))
```

### Question 6a
High glycosaminoglycans (GAGs) urine levels can progressively damage tissue,  
and high levels are indictitive of inherited disease.Using the `GAGurine` dataset  
from the MASS package, what is the distribution of GAG assay values?   
Check the answer by `cat(answer6a)`.

### Question 6b
What's the difference in GAG distribution between children under 2 and  
all other children?  
see `cat(hint6b)` for data transform.  
Check the answer by `cat(answer6)`.

\pagebreak  



## Boxplot 
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(movies[movies$year > 1990,],aes(x=factor(year),y=rating)) + 
  geom_boxplot()
```
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(movies[movies$year > 1990,],aes(x=factor(year),y=rating,fill=factor(Short))) + 
  geom_boxplot()
```
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(movies[movies$year > 1990,],
aes(x=cut(year,breaks=seq(from=1991,to=2005,by=5)),y=rating,fill=factor(Short))) + 
geom_boxplot()
```


### Question 7 
What is the distribution of GAG assay levels for each age?  
see `cat(hint7)` for data transform.  
Check the answer by `cat(answer7)`.  

\pagebreak  



## Themes


#### Adjust the background color, tick marks, etc
#### store what we have so far to a variable
```{r}
iris.plot = ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length,color=Species,shape=Species)) +
  geom_point(size=4) +
  geom_point(size=2,color="grey")+
  ggtitle("Measurement of Flowers")+
  xlab("width (cm)")+
  ylab("length (cm)")
```
#### plot stored variable
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
iris.plot
```

#### Change the background to black and white
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
iris.plot + theme_bw()
```

#### Other themes to consider:
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
iris.plot + theme_classic()
```
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
iris.plot + theme_minimal()
```

#### Customize the theme and specify features...
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
iris.plot +
  theme(plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        legend.key = element_rect(fill = "white"),
        legend.background = element_rect(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(colour = "lightgrey", linetype = "dotted"),
        axis.line = element_line(),
        axis.text = element_text(colour = "black"))
```


#### See http://docs.ggplot2.org/current/theme.html for more options

## Question 8
Make the ugliest plot you can. Bonus points for discordant color schemes. 
Check the answer by `cat(answer8)`.  

## Thank You!

\pagebreak  



# Bonus Material
Go through at your own pace, or use as a reference.


## Legend Positioning 

 what options do we have for the position of the legend?
 examples borrowed from: http://www.r-bloggers.com/ggplot2-cheatsheet-for-scatterplots/
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
g1 <- ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point(aes(color=factor(vs)))
```

#### move legend to the inside
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
g2 <- g1 + theme(legend.position=c(1,1),legend.justification=c(1,1))
```

#### move legend to the bottom 
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
g3 <- g1 + theme(legend.position = "bottom")
```

##### change labels
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
g4 <- g1 + scale_color_discrete(name ="Engine",
                                labels=c("V-engine", "Straight engine"))
```


```{r,fig.width=7,fig.height=7,cache.lazy=TRUE}
grid.arrange(g2, g3, g4, nrow=1)
```

\pagebreak  



## Facets


#### Let's just look at PG, R, and PG-13 rated movies...
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
mov.df = movies[as.character(movies$mpaa) %in% c("PG","R","PG-13"),]
mov.df$mpaa = mov.df$mpaa[,drop=TRUE]
ggplot(mov.df,aes(x=rating,fill=factor(Short))) +
  geom_density(alpha=I(0.3)) + theme_bw() + 
  facet_wrap(~mpaa)
```

#### seperate movies by decade, rating: view rating distribution
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
mov.df$decade = mov.df$year %/% 10 * 10
mov50plus.df = mov.df[mov.df$decade >= 1950,]
# facet_grid( y ~ x )
ggplot(mov50plus.df, aes(x=rating,fill=factor(Drama))) + 
  geom_density(alpha=I(0.4))+ theme_bw()+
  facet_grid(decade ~ mpaa )
```

\pagebreak  



##   Grids 

Set up two different plots of iris data for two species, and view them 
side by side. 
```{r,fig.width=7,fig.height=4,cache.lazy=TRUE}
irisSmall.df <- subset(iris, Species != "versicolor")

p1 <- ggplot(irisSmall.df, aes(x= Sepal.Width, y = Sepal.Length)) + geom_point(aes(color=factor(Species)))
p2 <- ggplot(irisSmall.df, aes(x = Petal.Width, y = Petal.Length)) + geom_point(aes(color=factor(Species)))

#' all plots in one row
grid.arrange(p1,p2,nrow=1)
```

\pagebreak  



##    Coordinates
#### coord_flip
flip the coords, easier to read x labels... 

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(mov50plus.df,aes(x=factor(year),y=length))+geom_boxplot()+theme_bw()
```

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(mov50plus.df,aes(x=factor(year),y=length))+geom_boxplot()+theme_bw() + coord_flip()
```

## coord_polar
How many teams do baseball players play for?
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
playerTeam.df = ddply(baseball,.(id),function(df)length(unique(df$team)))
teamsPerPlayer.df = as.data.frame(table(playerTeam.df$V1))

colnames(teamsPerPlayer.df) = c("teamsPlayedFor","numberOfPlayers")
teamsPerPlayer.df$teamsPlayedFor = as.numeric(teamsPerPlayer.df$teamsPlayedFor)
ggplot(teamsPerPlayer.df,aes(x=factor(1),y=numberOfPlayers,fill=factor(teamsPlayedFor))) +
  geom_bar(stat="identity") + 
  theme_bw() + 
  coord_polar(theta="y")
```

## coord_trans
apply a transform to your coordinates, simpler than applying it to the data...
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ct.plot = ggplot(diamonds, aes(carat,price))+geom_point(alpha=I(0.1),size=1)+
  theme_bw()
ct.plot
ct.plot + scale_y_log10()
ct.plot + scale_x_log10()
ct.plot + coord_trans(x="log10",y="log10")

#'# change the axis ticks w/ labels
ct.plot + scale_y_continuous(breaks= seq(0,20000,by=2500))+
  scale_x_continuous(breaks= seq(0,5,by=0.5))
```

##    Color/Fill

note: different scales are used for discrete and continuous data 
scale_color_manual for discrete data

```{r}
scm.plot = ggplot(iris,aes(x=Sepal.Length,y=Sepal.Width,color=factor(Species))) + geom_point(size=3)+
  theme_bw()
```

#### automatic assignment
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
scm.plot + scale_color_manual(values=c("black","red","blue")) 
```

#### specify colors and assignment to factor values
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
scm.plot + scale_color_manual(values=c("setosa"= "black","versicolor"= "red","virginica"= "blue"))
```

#### scale_colour_gradient for continuous data
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(baseball,aes(x=hr,y=rbi,color=h))+ geom_point()+
  theme_bw() + 
  scale_colour_gradient(low="red",high="green")
```

#### scale_colour_gradientn provides smoothing over multiple color values...
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
sin.df <-data.frame(x=1:1000,y=sin(1:1000),rr=1:1000 - (2*pi)*(1:1000 %/% 2*pi))
ggplot(sin.df,aes(x,y,color=rr))+geom_point(size=2)+theme_bw()+
  scale_colour_gradientn(colours=c("black","yellow","blue"))
```

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
scgn.plot = ggplot(baseball,aes(x=hr,y=rbi,color=h))+ geom_point()+ theme_bw()  
scgn.plot + scale_color_gradientn(colours=rainbow(7))
```

#### you can cycle through the colors for better resolution
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
scgn.plot + scale_color_gradientn(colours=rep(rainbow(10),2))
```

### scale_colour_brewer, offers a set of nice palettes for discrete coloring:
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(iris,aes(x=Sepal.Length,y=Sepal.Width,color=Species)) + 
  geom_point(size=3) + 
  scale_colour_brewer(palette="Set1")+theme_bw()
```

### scale_fill_* 
works the same as the scale_colour_*, and is used when the geom is "filled" instead of "colored"
(note this change from factor to integer is improper)
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}

teamsPerPlayer.df$teamsPlayedForInt = 1:12

ggplot(teamsPerPlayer.df,aes(x=factor(1),y=numberOfPlayers,fill=teamsPlayedForInt)) +
  geom_bar(stat="identity") +
  theme_bw() +
  coord_polar(theta="y")+
  scale_fill_gradientn(colours=rainbow(10))
```

#### we can try again using discrete values to get proper labels...
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}

ggplot(teamsPerPlayer.df,aes(x=factor(1),y=numberOfPlayers,fill=factor(teamsPlayedFor))) +
  geom_bar(stat="identity") +
  theme_bw() +
  coord_polar(theta="y")+
  scale_fill_manual(values=rainbow(12),guide= guide_legend(ncol = 4,byrow=TRUE))+
  theme(legend.position = "top",
        panel.grid.minor.y = element_line(colour = "grey"),
        panel.grid.major.y = element_line(colour = "grey"),
        axis.text.x = element_text(size=rel(1.5)),
        axis.text.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.y = element_blank()) + 
  scale_y_discrete(limits= seq(0,1228,by=100))+
  xlab("") 
```   

\pagebreak  



#  Advanced Examples


## ex. 1 
objective: look at the variance and mean of samples of discrete data
approach: create a function that initiates the dataset, then returns a
function whose arguments set the plot window
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}

makePlotFunction <-function(size,fun,mean,range,rangeLower=2){
  df<-as.data.frame(t(as.matrix(
    sapply(floor(runif(size)*(range))+ rangeLower,
           function(x){vec<-sapply(fun(x,mean),floor);
                       c(mean(vec),var(vec),length(vec))}))))#end of data.frame
  colnames(df)<-c("mean","variance","n");
  function(pMin,pMax,xVec,yVec){
    df$pSize = as.numeric(df$n)
    ggplot(df,aes(x=mean,y=variance,color=factor(n),size=pSize))+
      geom_point()+
      scale_size(range = c(pMin,pMax))+
      theme_bw()+
      xlim(xVec)+
      ylim(yVec)}}

# set the function
f1= makePlotFunction(size=10000,fun=rnorm,mean=2,range=80,rangeLower=100)
f1(xVec=c(1,2),yVec=c(0,2),pMin=2.2,pMax=0.2)
f1(xVec=c(1.4,1.6),yVec=c(1.1,1.4),pMin=2.2,pMax=0.2)

# Mosiac Plot
library(MASS)
data(caith)

# melt function :: wide data.frame -> long data.frame
caith.melt= melt(
  cbind(as.data.frame(caith),
        row=rownames(caith)),
  id.var=c("row"))

#' rename the columns
colnames(caith.melt) = c("eyeColor","hairColor","proportion")

ggplot(caith.melt ,aes(x=eyeColor,y=proportion,fill=hairColor,color=hairColor)) +
  geom_bar(position="fill",stat="identity",size=1) + 
  scale_fill_manual(values=c(topo.colors(4),topo.colors(1))) + 
  theme_bw() + 
  scale_color_manual(values=rep("black",6))
```   

## Mandelbrot set
```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}

normIter3 = function(n,z,P=2,N=200){n - log( (log(abs(z),base=N)),base=P) }
iterate.until.escape <- function(z, c, trans, cond, max=50, response=dwell) {
  #we iterate all active points in the same array operation,
  #and keeping track of which points are still iterating.
  active <- seq_along(z)
  dwell <- z
  dwell[] <- 0
  for (i in 1:max) {
    z[active] <- trans(z[active], c[active]);
    survived <- cond(z[active])
    dwell[active[!survived]] <- normIter3(i, z[active[!survived]])
    #dwell[active[!survived]] <- i
    active <- active[survived]
    if (length(active) == 0) break
  }
  eval(substitute(response))
}


getArrayZoom <- function(re.vec=c(-2,1),im.vec=c(-1.5,1.5),max=25,len=500){
  re = seq(re.vec[1], re.vec[2], len=len)
  im = seq(im.vec[1], im.vec[2], len=len)
  c <- outer(re, im, function(x,y) complex(real=x, imaginary=y))
  x <- iterate.until.escape(array(0, dim(c)), c,  
                            function(z,c)z^2+c, 
                            function(z)abs(z) <= 2,                
                            max=max) # was 100
  melt(x)
}

xRel = c(-2,0.5)
yImg = c(-1.5,1.5)

ggplot(getArrayZoom(re.vec=xRel,
                    im.vec=yImg),
       aes(Var1,Var2,fill=as.numeric(value))) +
  geom_raster() + 
  scale_fill_gradientn(colours=c("black",(rainbow(7)),na.value="black")) + 
  theme_minimal()+
  xlab("pixels of Real values")+
  ylab("imaginary values")
```

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(getArrayZoom(re.vec=c(-0.11,-.09),im.vec=c(0.95,0.965)), aes(Var1,Var2,fill=as.numeric(value)^(1/1))) + 
  geom_raster() + 
  scale_fill_gradientn(colours=rainbow(7),na.value="black") + 
  theme_minimal()
```


## Path 
instead of a line, we can follow each subsequent point to the next...

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
# create a random walk:
steps = 500
#' set our reduce function, so we don't have to use loops...
reduceFn = function(x,y)c(x[1]+runif(1)*2 - 1,x)
#' create a function that returns a data.frame with a random walk of "steps" steps
randomWalk  = function(steps)data.frame(x= Reduce(f=reduceFn,1:steps,init=c(1)),
                                        y=  Reduce(f=reduceFn,1:steps,init=c(1)),
                                        index=1:(steps + 1))

#' if you use a line plot, notice that the x values are ordered...
ggplot(randomWalk(1000),aes(x,y,color=index))+geom_line() 
```

```{r,fig.width=5,fig.height=4,cache.lazy=TRUE}
ggplot(randomWalk(1000),aes(x,y,color=index))+geom_path() + geom_point()
```







